---100 title: "Template" author: "Studentnames and studentnumbers here" date: "`r Sys.Date()`" output: pdf_document ---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(readxl)
```

# Title Page

Teammembers:\
Koen Damen (2889378)

Thomas Dop (2898269)

Tim de Marez Oyens (2882348)

Timo Bakker (2863621)

Daniel Ferrier ()

Clémence van Gils ()

Tutorial 5 Group 3

Tutorial lecturer: Simon Loop

Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 The GDP-Happiness Disconnect

Governments often use GDP as a measure of how well a society is doing. Data from the WHR (World Happiness Report) shows that a growth in GDP does not automatically mean an increase in the happiness of the citizens.\
\
While people in richer countries are generally happier compared to those in poorer countries, this correlation stops working at a certain point. Once a country has reached a certain amount of wealth, needed to secure basic needs like food, healthcare and housing, more economic growth does **not** make the people happier. This gap between the rising wealth and stagnant happiness is a big social problem. It proves that only focussing on economic growth can cause the leaders of a country to ignore deeper underlying human needs.\

Why this is relevant

- GDP blindsides policy makers\
  The GDP of a country can grow whilst its people face rising depression and loneliness. Relying only on financial metrics leaves governments blind to these social problems.

- The Easterlin Paradox\
  After basic needs are met, happiness depends on non-economic factors. Things like social support, public safety and freedom matter a lot more than extra wealth.

- Wealth inequality\
  GDP measures total wealth per citizen, but not how it is shared. If economic growth only benefits a small elite group while the cost of living for the majority of people rises, overall happiness will drop

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}


#data <- read_xlsx("/Users/timdemarezoyens/Documents/imf-dm-export-20260603-2.xlsx")
#saveRDS(data, "data/dataset")

dataset1 <- read.csv("data/dataset1.csv")
dataset2 <- read.csv("data/dataset2.csv")

library(tidyverse)

dataset <- read_csv("data/dataset1.csv")




```

.

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
