---100 title: "Template" author: "Studentnames and studentnumbers here" date: "`r Sys.Date()`" output: pdf_document ---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(readxl)
```

# Title Page

Teammembers:\
Koen Damen (2889378)

Thomas Dop (2898269)

Tim de Marez Oyens (2882348)

Timo Bakker (2863621)

Daniel Ferrier (2900731)

Clémence van Gils (2898031)

Tutorial 5 Group 3

Tutorial lecturer: Simon Loop

Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 The Relationship Between GDP Growth and Unemployment

GDP is one of the most important economic indicators. GDP measures the total value of all goods and services produced in a country . Governments often use GDP as a measure of how well a society is doing. If GDP falls, consumption and investment will decline, and people will lose their jobs. This will lead to a rise in the unemployment rate. Conversely, if GDP grows, the unemployment rate will actually fall, as more investments are made, consumption increases, and, as a result, more people will be hired. In addition, high unemployment entails high costs for the government.

That is why it is important to conduct research to gain insight into the relationship between unemployment and GDP. By examining this relationship, we can determine the extent to which economic growth contributes to increased employment. The findings could help policymakers to take measures that stimulate economic growth and also help to reduce unemployment. In this report we quantify the the relationship between the GDP per capita (*Gross Domestic Product (GDP) per Capita - Data Portal - United Nations Economic Commission for Europe*, n.d.) and the unemployment rate (*Unemployment Rate - Data Portal - United Nations Economic Commission for Europe*, n.d.) from 2005 until 2024.

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}


#dataset1 <- read.csv("/Users/bakkertje/Downloads/data_20260609135240.csv")

#write.csv(dataset1, "data/dataset1")

dataset1 <- read.csv("data/dataset1")

#dataset2 <- read.csv("/Users/bakkertje/Downloads/data_20260609135110.csv")

#write.csv(dataset2, "data/dataset2")

dataset2 <- read.csv("data/dataset2")

dataset1$X <- NULL
dataset1$Alpha3Code <- NULL
dataset1$CountryCode <- NULL
dataset1$IndicatorCode <- NULL
dataset1$X.1 <- NULL



dataset2$IndicatorCode <- NULL
dataset2$X.1 <- NULL
dataset2$CountryCode <- NULL
dataset2$Alpha3Code <- NULL
dataset2$X <- NULL




```

.

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

Dataset1 contains the procentual unemployment rate for 55 countries from 2005 to 2025.\
Dataset 2 contains the GDP per capita in USD adjusted by purchasing power parities. We have data for 56 countries for the years 2005 - 2024.\
For some countries we are missing some years of GDP and unemployment value.

```{r}
names(dataset1)
names(dataset2)
head(dataset1)
head(dataset2)

dim(dataset1)
dim(dataset2)

```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

\<\<\<\<\<\<\< Updated upstream

```{r data_cleaning}



dataset1_cleaned <- dataset1 %>%
  select(IndicatorName, CountryName, PeriodCode , Value) 


dataset2_cleaned <- dataset2 %>% 
  select(IndicatorName, CountryName, PeriodCode, Value)

```

## 3.2

Variable 1

```         
```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

*Gross domestic product (GDP) per capita - Data Portal - United Nations Economic Commission for Europe*. (n.d.-b). <https://w3.unece.org/PXWeb/en/Table?IndicatorCode=12>

*Unemployment rate - Data Portal - United Nations Economic Commission for Europe*. (n.d.). <https://w3.unece.org/PXWeb/en/Table?IndicatorCode=43>
