------------------------------------------------------------------------

title: "The relationship between GDP per Capita and Unemployment"

author: "Tutorial 5 group 3"

date: "`r Sys.Date()`"

output: pdf_document

------------------------------------------------------------------------

# Set-up your environment

```{r package_install, include=FALSE, eval=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("dplyr")
install.packages("ggplot2")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(readxl)
library(dplyr)
library(ggplot2)
```

# Title Page

Teammembers:\
Koen Damen (2889378)

Thomas Dop (2898269)

Tim de Marez Oyens (2882348)

Timo Bakker (2863621)

Daniel Ferrier (2900731)

Clémence van Gils (2898031)

Tutorial 5 Group 3

Tutorial lecturer: Simon Loop

# Part 1 - Identify a Social Problem

## 1.1 The Relationship Between GDP Growth and Unemployment

GDP is one of the most important economic indicators. It measures the total value of all goods and services produced within a country. Governments often use GDP as a measure of how well a society is doing.

According to Okun (1962), there is a negative relationship between economic growth and unemployment: if economic activity falls, there is less production and labor demand and then people will lose their jobs. This will lead to a rise in unemployment. Conversely, if economy expands and GDP grows, unemployment will fall, as investment increases, consumption rises and, as a result, more people are hired. (Blanchard et al., 2021) Furthermore, high unemployment entails high costs for the government.

That is why it is important to conduct research to gain insight into the relationship between unemployment and GDP. By examining this relationship, we can determine the extent to which economic growth contributes to increased employment. The findings could help policymakers to take measures that stimulate economic growth and also help to reduce unemployment. In this report we quantify the the relationship between the GDP per capita (*Gross Domestic Product (GDP) per Capita - Data Portal - United Nations Economic Commission for Europe*, n.d.) and the unemployment rate (*Unemployment Rate - Data Portal - United Nations Economic Commission for Europe*, n.d.) from 2005 until 2024.\

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}


#dataset1 <- read.csv("/Users/bakkertje/Downloads/data_20260609135240.csv")

#write.csv(dataset1, "data/dataset1")

dataset1 <- read.csv("data/dataset1")

#dataset2 <- read.csv("/Users/bakkertje/Downloads/data_20260609135110.csv")

#write.csv(dataset2, "data/dataset2")

dataset2 <- read.csv("data/dataset2")






```

## 2.2 Provide a short summary of the dataset(s)

Dataset 1 contains the procentual unemployment rate for 55 countries from 2005 to 2025. Dataset 2 contains the GDP per capita in USD adjusted by purchasing power parities. We have data for 56 countries for the years 2005 to 2024. For some countries we are missing some years of GDP and unemployment value.

```{r}
names(dataset1)
names(dataset2)
head(dataset1)
head(dataset2)

dim(dataset1)
dim(dataset2)

```

```{r}
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

The variables contain SES (socialeconomic status) information. GDP per capita reflects economic performance and unemployment rate reflects the situation in the labour market. The data are drawn from administrative and statistical sources that have been compiled and made available via the UNECE database.

**Dataset 1** 

Dataset 1 is a panel dataset with measurements from 55 countries from 2005 to 2025.\
The main variable is the unemployment rate, measured as a percentage of the total labour force. This is a continuous quantitative variable.\
Countries are identified by categorical variables such as CountryCode and Alpha3Code. The variable PeriodCode is the identifier for the time dimension, shown in years.\
The metadata variables, like Indicator Code and Variable name, are used to show what is being measured by set variables.

**Dataset 2** \
Dataset 2 is also a panel dataset, containing measurements from 56 countries from 2005 to 2024.\
The main variable is GDP per capita, measured in PPP-adjusted US dollars. This is a continuous quantitative variable.\
Similar to Dataset 1, countries are identified by categorical variables and the PeriodCode tracks the time dimension.\
The metadata variables, like Indicator Code and Variable name, are used to show what is being measured by set variables.

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

The column Unnamed: 9 is empty and should be removed during the data-cleaning process.

```{r data_cleaning}

dataset1_cleaned <- dataset1 %>%
  select(IndicatorName, CountryName, PeriodCode , Value) 


dataset2_cleaned <- dataset2 %>% 
  select(IndicatorName, CountryName, PeriodCode, Value)

merged_data <- inner_join(
  dataset1_cleaned,
  dataset2_cleaned,
  by = c("CountryName", "PeriodCode")
)

```

## 3.2

Variable 1

```{r}

library(dplyr)

lagged_gdp_table <- merged_data %>%
  arrange(CountryName, PeriodCode) %>%
  group_by(CountryName) %>%
  mutate(
    # Force Value.x to be numeric so the math works
    Value.x = as.numeric(Value.x),
    
    # Calculate GDP growth and its lag
    GDPGrowth = ((Value.x - lag(Value.x)) / lag(Value.x)) * 100,
    LaggedGDPGrowth = lag(GDPGrowth)
  ) %>%
  select(
    CountryName,
    PeriodCode,
    Value.x,
    GDPGrowth,
    LaggedGDPGrowth
  ) %>%
  ungroup()

head(lagged_gdp_table)

```

```         
```

Variable 2

```{r gen_var2}

library(dplyr)

elasticity_table <- merged_data %>%
  arrange(CountryName, PeriodCode) %>%
  group_by(CountryName) %>%
  mutate(
    # Ensure both columns are treated as numbers
    Value.x = as.numeric(Value.x),
    Value.y = as.numeric(Value.y),
    
    # Now run your calculations safely
    GDPGrowth = ((Value.y - lag(Value.y)) / lag(Value.y)) * 100,
    UnemploymentChange = Value.x - lag(Value.x),
    Elasticity = UnemploymentChange / GDPGrowth
  ) %>%
  ungroup()

head(elasticity_table)
```

## 3.3 Visualize temporal variation

```{r}

library(ggplot2)
library(dplyr)

plot_data <- elasticity_table %>%
  group_by(PeriodCode) %>%
  summarise(
    # CHANGE 'Value.y' to whatever your actual GDP column is named
    `Avg GDP per Capita` = mean(Value.y, na.rm = TRUE), 
    
    # CHANGE 'Value.x' to whatever your actual Unemployment column is named
    `Avg Unemployment Rate` = mean(Value.x, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(
    cols = c(`Avg GDP per Capita`, `Avg Unemployment Rate`), 
    names_to = "Metric", 
    values_to = "Value"
  )

# 2. Plot
ggplot(plot_data, aes(x = PeriodCode, y = Value, color = Metric)) +
  geom_line(size = 1) +
  geom_point() +
  facet_wrap(~Metric, ncol = 1, scales = "free_y") + 
  scale_x_continuous(breaks = seq(2004, 2024, by = 2)) + 
  labs(
    title = "Temporal Variation of GDP per Capita and Unemployment",
    x = "Year",
    y = "Average Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## 3.4 Visualize spatial variation

```{r visualise_map}
library(dplyr)
library(ggplot2)

merged_data$Value.x <- as.numeric(merged_data$Value.x)

spatial_data <- merged_data %>%
  group_by(CountryName) %>%
  summarise(
    Avg_Unemployment = mean(Value.x, na.rm = TRUE)
  ) %>%
  arrange(desc(Avg_Unemployment)) %>%
  slice(1:10)

ggplot(
  spatial_data,
  aes(
    x = reorder(CountryName, Avg_Unemployment),
    y = Avg_Unemployment
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Countries with the Highest Average Unemployment Rate",
    x = "Country",
    y = "Average Unemployment Rate (%)"
  ) +
  theme_minimal()


```

Here you provide a description of why the plot above is relevant to your specific social problem.

This figure shows the differences in unemployment rates between countries. We can see that unemployment is not the same everywhere. Countries such as North Macedonia, Bosnia and Herzegovina, and Georgia have much higher unemployment rates than other countries in the dataset. This is important for our social problem because high unemployment can lead to poverty, financial problems, and a lower quality of life. The graph shows that some countries are affected by unemployment much more than others.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r}

boxplot_data <- elasticity_table %>%
  filter(!is.na(Value.y), !is.na(Value.x)) %>%
  mutate(
    GDP_Group = ifelse(
      Value.y >= median(Value.y, na.rm = TRUE),
      "High GDP countries",
      "Low GDP countries"
    )
  )

ggplot(boxplot_data, aes(x = GDP_Group, y = Value.x)) +
  geom_boxplot() +
  labs(
    title = "Sub-population Variation: Unemployment in High- and Low-GDP Countries",
    x = "GDP per Capita Group",
    y = "Unemployment Rate (%)"
  ) +
  theme_minimal()
```

This boxplot compares the distribution of unemployment rates between low-GDP and high-GDP countries. The countries are grouped based on whether their GDP per capita is below or above the median GDP per capita in the dataset.

The main purpose of the graph is to see whether unemployment rates differ between richer and poorer countries. If the box for low-GDP countries is higher, it suggests that countries with lower GDP per capita generally have higher unemployment. If the high-GDP box is lower, this suggests that wealthier countries may have more stable labour markets.

The line inside each box shows the median unemployment rate. The box itself shows the middle 50% of the observations. The lines extending from the box show the wider range of values, while individual dots represent possible outliers.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
library(dplyr)
library(ggplot2)

event_data <- merged_data %>%
  mutate(
    Value.x = as.numeric(as.character(Value.x)),
    PeriodCode = as.numeric(as.character(PeriodCode))
  ) %>%
  group_by(PeriodCode) %>%
  summarise(
    Avg_Unemployment = mean(Value.x, na.rm = TRUE)
  ) %>%
  filter(PeriodCode >= 2006 & PeriodCode <= 2012)

ggplot(event_data, aes(x = PeriodCode, y = Avg_Unemployment)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = 2008, linetype = "dashed", color = "red") +
  annotate(
    "text",
    x = 2008,
    y = max(event_data$Avg_Unemployment, na.rm = TRUE),
    label = "2008 financial crisis",
    color = "red",
    vjust = -0.5
  ) +
  labs(
    title = "Unemployment Around the 2008 Financial Crisis",
    x = "Year",
    y = "Average Unemployment Rate (%)"
  ) +
  theme_minimal()
```

This figure shows the average unemployment rate around the 2008 financial crisis. After 2008, unemployment increased noticeably in the countries included in the dataset. This suggests that the crisis may have had a negative impact on employment. The figure is relevant because higher unemployment can lead to poverty, financial insecurity, and a lower quality of life.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Okun, A. M. (1962). Potential GNP: Its measurement and significance. *Proceedings of the Business and Economic Statistics Section of the American Statistical Association*, 98–104.

Blanchard, O., Amighini, A., & Giavazzi, F. (2021).*Macroeconomics: a European perspective*(4th ed.) [Macroeconomics]. Pearson Education Limited.<https://www.pearson.com/uk>

*Gross domestic product (GDP) per capita - Data Portal - United Nations Economic Commission for Europe*. (n.d.-b).<https://w3.unece.org/PXWeb/en/Table?IndicatorCode=12>

*Unemployment rate - Data Portal - United Nations Economic Commission for Europe*. (n.d.).<https://w3.unece.org/PXWeb/en/Table?IndicatorCode=43>\
\
