---100 title: "Template" author: "Studentnames and studentnumbers here" date: "`r Sys.Date()`" output: pdf_document ---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("dplyr")
install.packages("ggplot2")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(readxl)
library(dplyr)
library(ggplot2)
```

# Title Page

Teammembers:\
Koen Damen (2889378)

Thomas Dop (2898269)

Tim de Marez Oyens (2882348)

Timo Bakker (2863621)

Daniel Ferrier (2900731)

Clémence van Gils (2898031)

Tutorial 5 Group 3

Tutorial lecturer: Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 The Relationship Between GDP Growth and Unemployment

GDP is one of the most important economic indicators. GDP measures the total value of all goods and services produced within a country. Governments often use GDP as a measure of how well a society is doing. According to Okun (1962), there is a negative relationship between economic growth and unemployment: if GDP falls, consumption and investment will decline and people will lose their jobs. This will lead to a rise in unemployment. Conversely, if GDP grows, unemployment will fall, as investment increases, consumption rises and, as a result, more people are hired. (Blanchard et al., 2021) Furthermore, high unemployment entails high costs for the government.

That is why it is important to conduct research to gain insight into the relationship between unemployment and GDP. By examining this relationship, we can determine the extent to which economic growth contributes to increased employment. The findings could help policymakers to take measures that stimulate economic growth and also help to reduce unemployment. In this report we quantify the the relationship between the GDP per capita (*Gross Domestic Product (GDP) per Capita - Data Portal - United Nations Economic Commission for Europe*, n.d.) and the unemployment rate (*Unemployment Rate - Data Portal - United Nations Economic Commission for Europe*, n.d.) from 2005 until 2024.\
\
is gecheckt voor ai

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}


#dataset1 <- read.csv("/Users/bakkertje/Downloads/data_20260609135240.csv")

#write.csv(dataset1, "data/dataset1")

dataset1 <- read.csv("data/dataset1")

#dataset2 <- read.csv("/Users/bakkertje/Downloads/data_20260609135110.csv")

#write.csv(dataset2, "data/dataset2")

dataset2 <- read.csv("data/dataset2")






```

.

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

Dataset 1 contains the procentual unemployment rate for 55 countries from 2005 to 2025. Dataset 2 contains the GDP per capita in USD adjusted by purchasing power parities. We have data for 56 countries for the years 2005 to 2024. For some countries we are missing some years of GDP and unemployment value.

```{r}
names(dataset1)
names(dataset2)
head(dataset1)
head(dataset2)

dim(dataset1)
dim(dataset2)

```

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

**Data set 1**

Dataset1 is a panel dataset. It contains observations for 55 countries across years. The main variable is the unemployment rate, measured as the percentage of the labour force, shown as the variable Value. This is a continuous quantitative variable. Countries are identified by several variables as CountryCode, Alpha3Code, and CountryName. The time dimension is represented by PeriodCode. Dataset 1 also includes metadata variables like IndicatorCode and VariableName. These variables give information about what is being measured and how it is measured. 

#### **Data set 2**

Dataset 2 is also a panel dataset, containing observations for 56 countries across 20 time periods from 2005 - 2024. The main variable is GDP per capita, measured in PPP-adjusted US dollars. This is a continuous quantitative variable. Countries are identified in the same way as in Dataset1. The time dimension is represented by the variable PeriodCode. The dataset also includes metadata variables, such as IndicatorCode and IndicatorName. The latter defines what the variable Value is measuring, which is GDP per capita and is the main variable in this dataset. PeriodCode functions as the time variable. Most of these variables have been removed in the cleaning of the data.

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

The column Unnamed: 9 is empty and should be removed during the data-cleaning process.

\<\<\<\<\<\<\< Updated upstream

```{r data_cleaning}

dataset1_cleaned <- dataset1 %>%
  select(IndicatorName, CountryName, PeriodCode , Value) 


dataset2_cleaned <- dataset2 %>% 
  select(IndicatorName, CountryName, PeriodCode, Value)

merged_data <- inner_join(
  dataset1_cleaned,
  dataset2_cleaned,
  by = c("CountryName", "PeriodCode")
)

```

## 3.2

Variable 1

```{r}

library(dplyr)

lagged_gdp_table <- merged_data %>%
  arrange(CountryName, PeriodCode) %>%
  group_by(CountryName) %>%
  mutate(
    # Force Value.x to be numeric so the math works
    Value.x = as.numeric(Value.x),
    
    # Calculate GDP growth and its lag
    GDPGrowth = ((Value.x - lag(Value.x)) / lag(Value.x)) * 100,
    LaggedGDPGrowth = lag(GDPGrowth)
  ) %>%
  select(
    CountryName,
    PeriodCode,
    Value.x,
    GDPGrowth,
    LaggedGDPGrowth
  ) %>%
  ungroup()

head(lagged_gdp_table)

```

```         
```

Variable 2

```{r gen_var2}

library(dplyr)

elasticity_table <- merged_data %>%
  arrange(CountryName, PeriodCode) %>%
  group_by(CountryName) %>%
  mutate(
    # Ensure both columns are treated as numbers
    Value.x = as.numeric(Value.x),
    Value.y = as.numeric(Value.y),
    
    # Now run your calculations safely
    GDPGrowth = ((Value.y - lag(Value.y)) / lag(Value.y)) * 100,
    UnemploymentChange = Value.x - lag(Value.x),
    Elasticity = UnemploymentChange / GDPGrowth
  ) %>%
  ungroup()

head(elasticity_table)
```

## 3.3 Visualize temporal variation

```{r}
library(ggplot2)
library(dplyr)

plot_data <- elasticity_table %>%
  group_by(PeriodCode) %>%
  summarise(
    `Avg GDP per Capita` = mean(Value.y, na.rm = TRUE), 
    `Avg Unemployment Rate` = mean(Value.x, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(
    cols = c(`Avg GDP per Capita`, `Avg Unemployment Rate`), 
    names_to = "Metric", 
    values_to = "Value"
  )

# 2. Plot
ggplot(plot_data, aes(x = PeriodCode, y = Value, color = Metric)) +
  geom_line(size = 1) +
  geom_point() +
  facet_wrap(~Metric, ncol = 1, scales = "free_y") + 
  scale_x_continuous(breaks = seq(2004, 2024, by = 2)) + 
  labs(
    title = "Temporal Variation of GDP per Capita and Unemployment",
    x = "Year",
    y = "Average Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

## 3.4 Visualize spatial variation

```{r visualise_map}
library(dplyr)
library(ggplot2)

merged_data$Value.x <- as.numeric(merged_data$Value.x)

spatial_data <- merged_data %>%
  group_by(CountryName) %>%
  summarise(
    Avg_Unemployment = mean(Value.x, na.rm = TRUE)
  ) %>%
  arrange(desc(Avg_Unemployment)) %>%
  slice(1:10)

ggplot(
  spatial_data,
  aes(
    x = reorder(CountryName, Avg_Unemployment),
    y = Avg_Unemployment
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Countries with the Highest Average Unemployment Rate",
    x = "Country",
    y = "Average Unemployment Rate (%)"
  ) +
  theme_minimal()


```

Here you provide a description of why the plot above is relevant to your specific social problem.

This figure shows the differences in unemployment rates between countries. We can see that unemployment is not the same everywhere. Countries such as North Macedonia, Bosnia and Herzegovina, and Georgia have much higher unemployment rates than other countries in the dataset. This is important for our social problem because high unemployment can lead to poverty, financial problems, and a lower quality of life. The graph shows that some countries are affected by unemployment much more than others.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
library(ggplot2)

merged_data$Value.x <- as.numeric(as.character(merged_data$Value.x))
merged_data$Value.y <- as.numeric(as.character(merged_data$Value.y))

ggplot(
  merged_data,
  aes(
    x = log(Value.y),
    y = Value.x
  )
) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Relationship between GDP per Capita and Unemployment",
    x = "Log GDP per Capita",
    y = "Unemployment Rate (%)"
  ) +
  theme_minimal()
```

This figure shows the relationship between GDP per capita and unemployment rates. The downward trend line suggests a negative relationship between the two variables. In general, countries with a higher GDP per capita tend to have lower unemployment rates. This supports the idea that stronger economies often create more employment opportunities. However, the relationship is not perfect, which means that other factors also influence unemployment. This analysis is relevant because unemployment can contribute to poverty, financial insecurity, and social inequality.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list 

**Okun, A. M.** (1962). Potential GNP: Its measurement and significance. *Proceedings of the Business and Economic Statistics Section of the American Statistical Association*, 98–104.

Blanchard, O., Amighini, A., & Giavazzi, F. (2021). *Macroeconomics: a European perspective* (4th ed.) [Macroeconomics]. Pearson Education Limited. <https://www.pearson.com/uk>

*Gross domestic product (GDP) per capita - Data Portal - United Nations Economic Commission for Europe*. (n.d.-b). <https://w3.unece.org/PXWeb/en/Table?IndicatorCode=12>

*Unemployment rate - Data Portal - United Nations Economic Commission for Europe*. (n.d.). <https://w3.unece.org/PXWeb/en/Table?IndicatorCode=43>\
\
